---
title: "Forecasting San Diego Power Consumption"
author: "Hunter Blum, Mackenzie Carter, Saba Alemayehu"
date: '2022-11-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries
```{r}
library(readxl)
library(rstudioapi)
library(parsedate)
library(lubridate)
library(psych)
library(corrplot)
library(outliers)
library(ggpmisc)
library(gridExtra)
library(zoo)
library(forecast)

# Note - Tidyverse is a collection of packages, see the Attaching packages section below. Usually best to load last so its functions will mask over other packages.
library(tidyverse)
```

Set up cores for  models - You may need to install/update Java.
```{r}
library(parallel)
library(doParallel)

cluster <- makeCluster(detectCores() - 1)
registerDoParallel(cluster)
```

# EDA/Cleaning 
Read in data (So you can skip first part)
```{r}
setwd(dirname(getActiveDocumentContext()$path))
SD <- read.csv("SD.csv")
SD <- SD[,-1]
SD <- SD %>% select(Date, SDGE, starts_with("Hourly"))
head(SD)
```

## Structual and General EDA

First, lets take a broad look at our data, including variable types, descriptive statistics and NA counts. 

```{r}
str(SD)
describe(SD)
```

Evaluate missing values by column
```{r}
missing.values <- SD %>%
    gather(key = "key", value = "val") %>%
    mutate(is.missing = is.na(val)) %>%
    group_by(key, is.missing) %>%
    summarise(num.missing = n()) %>%
    filter(is.missing==T) %>%
    select(-is.missing) %>%
    arrange(desc(num.missing)) 

missing.values %>%
  ggplot() + 
    geom_bar(aes(x=reorder(key, num.missing), y=num.missing), stat = 'identity') +
    labs(x='Variable', y="Number of missing values", title='Figure X.x: Number of missing values') +
    coord_flip()
```

Evaluate NAs by observation - 304 observations are missing all weather data, we will either need to impute or stick with regression models. 
```{r}
NA_row <- rowSums(is.na(SD))
obs <- seq(1:nrow(SD))

NA_row <- data.frame(obs, NA_row)
NA_row %>% filter(NA_row == 16) %>% tally()
```

Properly format variables
```{r}
# Numeric
SD$HourlyDryBulbTemperature <- as.numeric(as.character(SD$HourlyDryBulbTemperature))
SD$HourlyDewPointTemperature <- as.numeric(as.character(SD$HourlyDewPointTemperature))
SD$HourlyPrecipitation <- as.numeric(as.character(SD$HourlyPrecipitation))
SD$HourlyVisibility <- as.numeric(SD$HourlyVisibility)
SD$HourlyWindDirection <- as.numeric(SD$HourlyWindDirection)

# Factors
SD$HourlyPresentWeatherType <- as.factor(SD$HourlyPresentWeatherType)
SD$HourlySkyConditions <- as.factor(SD$HourlySkyConditions)

# Only run this if you loaded in SD data from EDA
SD$Date <- as_datetime(SD$Date)
```
## Imputation
Since we're doing a time-based imputation, we won't separate training/testing data before imputing in case the first validation observation is missing values. 
```{r}
# NAs in wind gusts should be zero
SD <- SD %>% mutate(HourlyWindGustSpeed = ifelse(is.na(HourlyWindGustSpeed), 0, HourlyWindGustSpeed))

# Fill the rest of NAs with the last non-NA value
SD <- na.locf(SD)
```


## Numeric EDA 

```{r}
num <- SD %>% dplyr::select(where(is.numeric))
describe(num)
summary(num)
```

### Outliers 
```{r}
tests = lapply(num, grubbs.test) 
tests
```

The grubbs test shows our max value for wind speed, 33, is an outlier. This makes sense because wind speed ranges in the single digits- it is likely that 33 was a typo. 

Not sure 33 mph winds is a typo, could easily happen in a storm. 

### Correlation
```{r}
M = cor(num)
corrplot(M, addCoef.col = 'black')
```
As we can see from our corrplot, station pressure, altimeter setting, & sea level pressure are 100% correlated. Hourly dry bulb, wet bulb, and dew point temperatures are closely correlated as well. We will be keeping wet bulb temperature as that is analogous to what humans feel. Altimeter setting is another measure of pressure, so we will be keeping hourly station pressure for ease of understanding. 

Correlations below look much better
```{r}
drop2 <- c("HourlyDryBulbTemperature","HourlyAltimeterSetting", "HourlyDewPointTemperature", "HourlySeaLevelPressure")
num2 = num[,!(names(num) %in% drop2)]
M2 = cor(num2)
corrplot(M2, addCoef.col = 'black')
```

```{r}
ggplot(data = num2, aes(x= HourlyWetBulbTemperature, y=SDGE)) + 
  geom_point() + 
  stat_poly_line() + stat_poly_eq() +
  labs(title="SDGE Usage By Temp", 
         x="Wet Bulb Temp (°F)", y = "Hourly Energy Use (MWh)") 
  
```



Scatter plots for numeric features compared to hourly energy use - may need to explore non-linear options. 
```{r}
ScatPlotter.SD <- function(numvar){
  ggplot(num2, aes(x = num2[,numvar], y = SDGE)) +
    geom_point(shape = 1 , alpha = 0.2) +
    stat_poly_line() + stat_poly_eq(geom = "label") +
    xlab(colnames(num2[numvar]))
}

num_cols <- c(2:11)
Scatterplots <- lapply(num_cols, ScatPlotter.SD)
grid.arrange(grobs = Scatterplots, top = "Figure X.x")
```

Time Series for Numeric Features 
Will make adjustments/sort out features to make graphs easier to read
```{r}
num_wdate <- num2
num_wdate$Date <- SD$Date

TS_plotter <- function(x){
  ggplot(num_wdate, aes(x = Date)) +
  geom_line(aes(y = num_wdate[,x]))+
  ylab(colnames(num_wdate[x]))
}

PHP_cols <- c(3,4,5,6)
TWV_cols <- c(2,8,9,10,11)
TS_target <- TS_plotter(1) + ggtitle("Figure X.x: Time Series for SDGE") 
TS_PHP <- lapply(PHP_cols, TS_plotter)
TS_TWV <- lapply(TWV_cols, TS_plotter)

plot(TS_target) 
grid.arrange(grobs = TS_PHP, top = "Figure X.x: Time Series for Pressure, Humidity, and Precipitation Variables")
grid.arrange(grobs = TS_TWV, top = "Figure X.x: Time Series for Temperature, Wind, and Visibility Variables")
```

Distributions - also will make this look better tomorrow.
```{r}
Hist_plotter <- function(x){
  ggplot(num_wdate, aes(x = num_wdate[,x])) +
    geom_histogram() +
    xlab(colnames(num_wdate[x]))
}

Hist_target <- Hist_plotter(1) + ggtitle("Figure X.x: Distribution for SDGE") 
Hist_PHP <- lapply(PHP_cols, Hist_plotter)
Hist_TWV <- lapply(TWV_cols, Hist_plotter)

plot(Hist_target) 
grid.arrange(grobs = Hist_PHP, top = "Figure X.x: Distribution for Pressure, Humidity, and Precipitation Variables")
grid.arrange(grobs = Hist_TWV, top = "Figure X.x: Distribution for Temperature, Wind, and Visibility Variables")
```

## Categorical EDA

Category Count Tables - Bar plots were too messy since we have so many categories
```{r}
SD %>% group_by(HourlyPresentWeatherType) %>% tally() %>% arrange(desc(n))
SD %>% group_by(HourlySkyConditions) %>% tally() %>% arrange(desc(n))
```

ANOVAs to see if category affects SDGE - 
Wouldn't work with SkyConditions since there are 8,145 categories. I think that having so many groups is forcing a significant result for WeatherType, looking at our Tukey most categories aren't different, so we'll just stick with our numeric features. 
```{r}
WT_aov <- aov(SDGE ~ HourlyPresentWeatherType, data = SD)

summary(WT_aov)

WT_tukey <- TukeyHSD(WT_aov)
```

## Target EDA

We've already seen yearly seasonality in SDGE from time series above, lets evaluate a few days to see if there is also a daily seasonality. We'll evaluate the first couple days in Jan, Apr, Jul, and Oct. We definitely see a daily-based seasonality. So, we'll have to account for seasonality across the year and the day when modelling. 
```{r}
SD %>% filter(Date < '2019-01-03 18:00:00') %>% ggplot(aes(x = Date, y = SDGE)) + geom_line() + ggtitle("Figure X.x: Seasonality in January")
SD %>% filter(Date < '2019-04-03 18:00:00') %>% filter(Date > '2019-03-31 18:00:00') %>% ggplot(aes(x = Date, y = SDGE)) + geom_line() + ggtitle("Figure X.x: Seasonality in April")
SD %>% filter(Date < '2019-07-03 18:00:00') %>% filter(Date > '2019-06-30 18:00:00') %>% ggplot(aes(x = Date, y = SDGE)) + geom_line() + ggtitle("Figure X.x: Seasonality in July")
SD %>% filter(Date < '2019-10-03 18:00:00') %>% filter(Date > '2019-09-30 18:00:00') %>% ggplot(aes(x = Date, y = SDGE)) + geom_line() + ggtitle("Figure X.x: Seasonality in October")
```

# Modelling 

When incorporating weather variables into the model, we will lag the predictors behind by 1-hour. This will make it so the SDGE has approximately 1 hour and 5 minutes to take any action on the with a forecast (since weather originally came in at minute 54 of each hour + 1 minute to run the model and return the prediction).

To simulate a real-life scenario, we will use a sliding train/test window for models that need specific training periods for weights (like exponential smoothing/regression). So, the validation set will begin in the last month (Sep 2022), then we will slide 1 hour forward and rerun the model, iterating until we reach the end of the month. 

Any non-weighted method (Naive/moving average) will be calculated on the entire data set and then split for evaluation, since we can assume we'll always have the last n-observations we need for the forecast.

## Baseline Model - Naive Forecast

Create forecasts and calculate metrics
```{r}
# SD got knocked out of order somewhere, rearrange to make sure lagging works
SD <- SD %>% arrange(Date)

# Create Naive Forecasts
SD$Naive <- lag(SD$SDGE)

# Calculate Error
SD$Naive_err <- SD$SDGE - SD$Naive

# Split for evaluation
SD_tr <- SD %>% filter(Date < '2022-08-31 19:00:00') %>% na.omit()
SD_val <- SD %>% filter(Date > '2022-08-31 18:00:00')

# Calculate Metrics
Naive_rmse_tr <- sqrt(mean(SD_tr$Naive_err^2))
Naive_rmse_val <- sqrt(mean(SD_val$Naive_err^2))

paste0("We had a RMSE of ", round(Naive_rmse_val,2), " with naive forecasts on the validation data.")
```

Plot on validation data
```{r}
ggplot(SD_val, aes(x = Date)) +
  geom_line(aes(y = Naive), color = 'green', linewidth = 1) +
  geom_line(aes(y = SDGE), color = 'blue', alpha = 0.4, linewidth = 1) +
  ggtitle("Figure X.x: Naive Forecasts on Validation Data")

ggplot(SD_val, aes(x = Date, y = Naive_err)) +
  geom_line() + 
  ggtitle("Figure X.x: Residuals for Naive Forecasts on Validation Data")
```

## Moving Average Forecasts

We'll use a function to evaluate different window sizes
```{r}
SD_madf <- SD
RMSE_tr <- c()
RMSE_val <- c()

SD.ma <- function(k){
  
  # Calculate forecasts
  DailyMA <- rollmean(SD_madf$SDGE, k = k)
  
  # Add NAs to forecast vector so we can bind it to our df
  DailyMA <- c(rep(NA, k-1), DailyMA)
  
  SD_madf$DailyMA <- DailyMA
  
  # Calculate Error
  SD_madf$DailyMA_err <- SD_madf$SDGE - SD_madf$DailyMA
  
  # Split for eval
  SD_tr <- SD_madf %>% filter(Date < '2022-08-31 19:00:00') %>% na.omit()
  SD_val <- SD_madf %>% filter(Date > '2022-08-31 18:00:00')
  
  # Calculate Metrics
  RMSE_tr <- sqrt(mean(SD_tr$DailyMA_err^2))
  RMSE_val <- sqrt(mean(SD_val$DailyMA_err^2))
  
  # List and return
  RMSE_list <- list(k, RMSE_tr, RMSE_val)

}

k = c(1:24, seq(48, 168, 24))

# Create results data frame
SD_ma_res <- lapply(k, SD.ma)
SD_ma_res <- as.data.frame(do.call(rbind, SD_ma_res))
colnames(SD_ma_res) <- c("k", "RMSE_tr", "RMSE_val")

SD_ma_res <- SD_ma_res %>% filter(RMSE_tr > 0) 
best_window <- SD_ma_res[which.min(SD_ma_res$RMSE_tr), ]
best_window$k

MA_rmse_tr <- best_window$RMSE_tr[[1]]
MA_rmse_val <- best_window$RMSE_val[[1]]
```
Plot
```{r}
DailyMA_2 <- rollmean(SD_madf$SDGE, k = 2)
DailyMA_2 <- c(rep(NA, 1), DailyMA_2)


SD_madf$DailyMA_2 <- DailyMA_2

ggplot(SD_madf, aes(x = Date)) +
  geom_line(aes(y = SDGE), color = "blue", linewidth = 1, alpha = 0.4) +
  geom_line(aes(y = DailyMA_2), color = "darkgreen", linewidth = 1, alpha = 0.5)
```

## Auto-ARIMA with predictors

Split predictors and target - WetBulbTemp and StationPressure were the only variables with slight relationships, so we'll use those.
```{r}
X_tr <- SD_tr %>% select(HourlyWetBulbTemperature, HourlyStationPressure) %>% as.matrix()
X_val <- SD_val %>% select(HourlyWetBulbTemperature, HourlyStationPressure) %>% as.matrix()
y_tr <- SD_tr$SDGE
y_test <- SD_val$SDGE
```

Create Model
```{r}
auto_arima_model <- auto.arima(y_tr, xreg = X_tr, seasonal = TRUE, stepwise = FALSE)
summary(auto_arima_model)

auto_arima_rmse_tr <- sqrt(mean(auto_arima_model$residuals^2))

checkresiduals(auto_arima_model)
```

Forecast
```{r}
val_forecast <- forecast(auto_arima_model, xreg = X_val)

# Calculate errors
val_forecast_err <- SD_val$SDGE - val_forecast$mean
auto_arima_rmse_val <- sqrt(mean(val_forecast_err^2))
```

Plot
```{r}
autoplot(val_forecast, series = "Forecast") +
  coord_cartesian(xlim = c(38000, 39488))
```

## ETS
Create time series - Will just do it for everything in case we need predictor ts later
```{r}
X_tr_ts <- ts(X_tr, frequency = 24)
X_val_ts <- ts(X_val, frequency = 24)
y_val_ts <- ts(y_test, frequency = 24)
y_tr_ts <- ts(y_tr, frequency = 24)
```

Fit ETS
```{r}
SD_ets <- ets(y_tr_ts, model = "ZNA")
summary(SD_ets)
```

Forecast, Calculate RMSE, and Plot
```{r}
ets_forecast <- forecast(SD_ets, h=length(y_test))

#RMSE
val_forecast_err <- SD_val$SDGE - ets_forecast$mean
ets_rmse_val <- sqrt(mean(val_forecast_err^2))
ets_rmse_tr <- sqrt(mean(SD_ets$residuals^2))

#plot
autoplot(ets_forecast, series = "Forecast") +
  coord_cartesian(xlim = c(1550, 1650))
```
## Regression 
Based on our time series, it looks like we have additive seasonality with no trend. 

```{r}
SD_reg_addseason <- tslm(y_tr_ts ~ season)
SD_reg_addseason_pred <- forecast(SD_reg_addseason, h = length(y_val_ts))
```

Forecast, Calculate RMSE, and Plot
```{r}
reg_forecast <- forecast(SD_reg_addseason, h=length(y_test))

#RMSE
val_forecast_err <- SD_val$SDGE - reg_forecast$mean
reg_rmse_val <- sqrt(mean(val_forecast_err^2))
reg_rmse_tr <- sqrt(mean(SD_reg_addseason$residuals^2))

#plot
autoplot(reg_forecast, series = "Forecast") +
  coord_cartesian(xlim = c(1550, 1650))
```

## Manual ARIMA
Twice differenced ACF plot
```{r}
acf(diff(diff(X_tr)))
```

Twice differenced PACF plot
```{r}
#take a look pacf plot with differenced
pacf(diff(diff(X_tr)))
```

Model
```{r}
#Arima model
manual_arima_model<- Arima(y_tr, order=c(3,2,2),xreg=X_tr)
summary(manual_arima_model)
```

Residual Plot
```{r}
#check the residuals
checkresiduals(manual_arima_model)
```

Forecast, Calculate RMSE, and Plot
```{r}
manual_forecast<-forecast(manual_arima_model, xreg = X_val)

#RMSE
val_forecast_err <- SD_val$SDGE - manual_forecast$mean
manual_arima_rmse_val <- sqrt(mean(val_forecast_err^2))
manual_arima_rmse_tr <- sqrt(mean(manual_arima_model$residuals^2))

#plot
autoplot(manual_forecast, series = "Forecast") +
  coord_cartesian(xlim = c(38000, 39488))
```

## Manual ARIMA with Seasonality
Create daily ts for easier seasonal eval
```{r}
SD_ts <- ts(SD$SDGE, frequency = 24)
```

```{r}
Acf(SD_ts, lag.max = 72)
SD_ts %>% diff(lag = 24)  %>% Acf()
```
pacf
```{r}
Pacf(SD_ts)
SD_ts %>% diff(lag = 24) %>% Pacf()
```

Fit model - Acf shows strong MA(1) for non-seasonal and seasonal on twice differenced chart, same story with Pacf for our AR(p) component. We took a non-seasonal and seasonal difference so we'll train a ARIMA(1,1,1)(1,1,1) model
```{r}
seasonal_arima <- Arima(y_tr, order=c(1,1,1), seasonal = c(1,1,1), xreg=X_tr)
summary(seasonal_arima)
seasonal_arima %>% residuals() %>% ggtsdisplay()
```

Continue to tune - Definitely didn't include enough components, we'll keep playing around until we find good values. (3,1,3) looks good for non-seasonal components. 
```{r}
seasonal_arima_tuned <- Arima(y_tr, order=c(3,1,3), seasonal = c(2,1,2), xreg=X_tr)
summary(seasonal_arima_tuned)
seasonal_arima_tuned %>% residuals() %>% ggtsdisplay()
```

Forecast, Calculate RMSE, and Plot
```{r}
seasonal_forecast <- forecast(seasonal_arima_tuned, xreg = X_val)

#RMSE
val_forecast_err <- SD_val$SDGE - seasonal_forecast$mean
seasonal_arima_rmse_val <- sqrt(mean(val_forecast_err^2))
seasonal_arima_rmse_tr <- sqrt(mean(seasonal_arima_tuned$residuals^2))

#plot
autoplot(seasonal_forecast, series = "Forecast") +
  coord_cartesian(xlim = c(38000, 39488))
```

## Neural Network
```{r}
#size= (p+p+1)/2 = (3+2+1)/2 =3
SD_NN <- nnetar(y_tr_ts, p=3, q=2, size =3, xreg = X_tr)
```

```{r}
#prediction
SD_NN.pred<-forecast(SD_NN, xreg = X_val)
summary(SD_NN.pred)
```
```{r}
# Plot the errors for the training period
plot(SD_NN.pred$residuals, main = "Residual Plot for Training Period")

#Accuracy & Plot
accuracy(SD_NN.pred, y_test)
autoplot(SD_NN.pred, series = "Forecast") +
  coord_cartesian(xlim = c(1000, 2000))
```
Get metrics
```{r}
nnetar_rmse_tr <- sqrt(mean(SD_NN$residuals^2, na.rm = TRUE))
val_forecast_err <- SD_val$SDGE - SD_NN.pred$mean
nnetar_rmse_val <- sqrt(mean(val_forecast_err^2))
```

## avNNet
```{r}
modelFit <- avNNet(X_tr,
                   y_tr,
                   size = 5,
                   linout = TRUE,
                   trace = FALSE)
summary(modelFit)

nnpred1 <- predict(modelFit, X_val)
```


```{r}
accuracy(nnpred1, y_test)
```


# Results
Create Results df
```{r}
Model <- c("Naive", "2-Hour MA", "Auto ARIMA", "Manual ARIMA", "Seasonal ARIMA", "ETS", "Seasonal Reg.", "Neural Net.")
Tr_RMSE <- c(Naive_rmse_tr, MA_rmse_tr, auto_arima_rmse_tr, manual_arima_rmse_tr, seasonal_arima_rmse_tr, ets_rmse_tr, reg_rmse_tr, nnetar_rmse_tr)
Val_RMSE <- c(Naive_rmse_val, MA_rmse_val, auto_arima_rmse_val, manual_arima_rmse_val, seasonal_arima_rmse_val, ets_rmse_val, reg_rmse_val, nnetar_rmse_val)

Results_df <- data.frame(Model, Tr_RMSE, Val_RMSE)

Results_df <- Results_df %>% mutate(BaselineComp_tr = ifelse(Tr_RMSE == Naive_rmse_tr, "Baseline", ifelse(Tr_RMSE < Naive_rmse_tr, "Better", "Worse")))
Results_df <- Results_df %>% mutate(BaselineComp_val = ifelse(Val_RMSE == Naive_rmse_val, "Baseline", ifelse(Val_RMSE < Naive_rmse_val, "Better", "Worse")))

Results_df <- Results_df %>% mutate_if(is.character, as.factor)
```

Plot Training RMSE's
```{r}
ggplot(Results_df, aes(x = reorder(Model, -Tr_RMSE), y = Tr_RMSE, color = BaselineComp_tr)) + 
  geom_segment(aes(x=reorder(Model, -Tr_RMSE), xend = reorder(Model, -Tr_RMSE), y = 0, yend = Tr_RMSE)) +
  scale_color_manual(values = c("dodgerblue2", "green3", "firebrick2")) +
  geom_point(size = 9) + coord_flip() + labs(color = "Comparison to Baseline") +
  ylab("RMSE for Training Data (MWh)") + ggtitle("Figure X.x: Model Results for Training Data") + xlab("Model") + geom_text(aes(label = round(Tr_RMSE)), color = "black", size = 2.5, fontface = "bold")
```

Plot Validation RMSE's - Reordering isn't working when I flip coordinates. Need to find workaround.
```{r}
ggplot(Results_df, aes(x = reorder(Model, -Val_RMSE), y = Val_RMSE, color = BaselineComp_val)) + 
  geom_segment(aes(x=reorder(Model, -Val_RMSE), xend = reorder(Model, -Val_RMSE), y = 0, yend = Val_RMSE)) +
  scale_color_manual(values = c("dodgerblue2", "green3", "firebrick2")) +
  geom_point(size = 9) + coord_flip() + labs(color = "Comparison to Baseline") +
  ylab("RMSE for Validation Data (MWh)") + ggtitle("Figure X.x: Model Results for Validation Data") + xlab("Model") + geom_text(aes(label = round(Val_RMSE)), color = "black", size = 2.5, fontface = "bold")
```

